Coherence of Molecular Mechanisms In Major Human Disease and Traits

### 2020-02-25 manuscript/submission.BMC_Bioinformatics/Revision1/manuscript_BMC_Bioinformatics_R1.Rmd
Make repository for revised files

@@

We estimated coherence of 49 traits from five categories (Figure 3, Additional File 3, Additional File 5, 6). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Additional File 1). Several notable examples of low coherence include Obesity-related traits, Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

<!--
+-------+
| ![](Figures/Fig3_summary_traits.png)      |
| **Figure 3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).   |
+-------+
-->

@@

We further estimated coherence of 67 diseases from seven categories (Figure 4, Additional File 7, 8, Additional File 3). Metabolic diseases had high overall coherence (average coherence 0.91, Table 1), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively).

@@

<!--
+-------+
| ![](Figures/Fig4_summary_diseases.png)      |
| **Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).   |
+-------+
-->

@@

<!--New start-->
## Coherence captures distinct network properties than modularity

@@

Although our results did not show an association between coherence and modularity, our study is limited by using a single most common modularity measure [@Clauset_2004]. Although numerous methods for module detection have been described [@Chen:2008aa; @Dobrin:2009aa; @Jia:2011aa; @Agrawal:2018aa], comparing modules between diseases is plagued by normalization issues described in the Results section. Our future direction includes the development of a strategy for normalization of modularity by network size and comparing it with normalized coherence.

## Modularity estimates

@@

Although the normalized coherence derived using different PPI databases can be compared among phenotypes, it tells nothing whether a given coherence estimate is significantly different from random. We developed a permutation test aimed to provide a measure of the stability of the coherence. Intuitively, the permutation test defines how different the coherence of a given phenotype from that of random networks of the same size. It provides additional protection against cases where coherence may appear high, yet, it may be a consequence of small network size providing unstable degree distributions. Testing the coherence estimates for significance helps against conclusions of the overall low coherence (e.g., anthropometric traits, Figure 3), when the coherence of individual traits is high (e.g., individual anthropometric traits, Figure 5B). Coupled with the permutation test, our coherence estimation provides a robust framework for quantifying the coherence estimates of phenotype-associated genes, advancing our understanding of the corresponding molecular mechanisms.-->

@@

<!--For each disease, we find the absolute value of the difference between its slope and the random coherence slope. Next, we combine the genes for each disease with those of the random diseases and take two random samples of equal size to the disease gene set. We then calculate the absolute difference in slopes between the randomly sampled gene sets. This process is repeated 1000 times. Finally, we obtain a permutation p-value by taking the number of times the difference in slopes between random samples were less than the overall difference in slopes and dividing by 1000. -->

@@

<!--
# Supplementary material

@@

**Figure 3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).

**Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).

>>>>

<!-- New in Revision 1 -->
## Networks with high/low coherence have distinct degree distributions

Our measure of network coherence is expected to be proportional to the level of internal and/or external degree distributions. To investigate the relationship between coherence and degree distributions, we selected phenotype networks with low and high coherence. The thresholds for low and high normalized coherence were selected as <0.58 and >0.85, corresponding to the first and third quartiles of the coherence range, respectively. The Biogrid-generated networks were omitted because they had insufficient number of genes to form well-defined degree distributions. Expectedly, the internal degree distributions for high- and low-coherence networks were significantly different from random (Bonferroni-corrected two-tailed Wilcoxon p-values < 5.30E-12), and these results were consistent when using different PPI databases (Figure 3, Additional File 9). Similarly, the internal degree distribution for high-coherence networks was significantly larger than that of low-coherence networks (p-value = 2.01E-16/2.12E-4 for STRING/STRING filtered databases, respectively). Similar observations were observed for external degree distributions of high- and low-coherence networks (Figure 3, Additional File 9). In summary, these results support the notion that networks of high coherence have significantly more internal and external interactions as compared with low-coherence and random networks.
<!--
+-------+
| ![](Figures/Fig3_degree_distributions.png)      |
| **Figure 3. Degree distributions differ for high/low-coherence networks.** Internal and external degree distributions for high- and low-coherence networks, and size-matched random networks. STRING and STRING Filtered indicate the corresponding PPI database. Bonferroni-corrected Wilcoxon p-values comparing the distributions are in Additional File 9.  |
+-------+
-->
<!-- End New in Revision 1 -->

@@

We estimated coherence of 49 traits from five categories (Figure 4, Additional File 3, Additional File 5, 6). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Additional File 1). Several notable examples of low coherence include Obesity-related traits, Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

<!--
+-------+
| ![](Figures/Fig3_summary_traits.png)      |
| **Figure 4. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).   |
+-------+
-->

@@

We further estimated coherence of 67 diseases from seven categories (Figure 5, Additional File 7, 8, Additional File 3). Metabolic diseases had high overall coherence (average coherence 0.91, Table 1), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively).

@@

<!--
+-------+
| ![](Figures/Fig4_summary_diseases.png)      |
| **Figure 5. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).   |
+-------+
-->

@@

<!-- New in Revision 1 -->
## Coherence captures distinct network properties than modularity

@@

Although our results did not show an association between coherence and modularity, our study is limited by using a single most common modularity measure [@Clauset_2004]. Although numerous methods for module detection have been described [@Chen:2008aa; @Dobrin:2009aa; @Jia:2011aa; @Agrawal:2018aa], comparing modules between diseases is plagued by normalization issues described in the Results section. Our future direction includes the development of a strategy for normalization of modularity by network size and comparing it with normalized coherence.
<!-- End New in Revision 1 -->

@@

Although the normalized coherence derived using different PPI databases can be compared among phenotypes, it tells nothing whether a given coherence estimate is significantly different from random. We developed a permutation test aimed to provide a measure of the stability of the coherence. Intuitively, the permutation test defines how different the coherence of a given phenotype from that of random networks of the same size. It provides additional protection against cases where coherence may appear high, yet, it may be a consequence of small network size providing unstable degree distributions. Testing the coherence estimates for significance helps against conclusions of the overall low coherence (e.g., anthropometric traits, Figure 4), when the coherence of individual traits is high (e.g., individual anthropometric traits, Figure 5B). Coupled with the permutation test, our coherence estimation provides a robust framework for quantifying the coherence estimates of phenotype-associated genes, advancing our understanding of the corresponding molecular mechanisms.-->

@@

<!--For each disease, we find the absolute value of the difference between its slope and the random coherence slope. Next, we combine the genes for each disease with those of the random diseases and take two random samples of equal size to the disease gene set. We then calculate the absolute difference in slopes between the randomly sampled gene sets. This process is repeated 1000 times. Finally, we obtain a permutation p-value by taking the number of times the difference in slopes between random samples were less than the overall difference in slopes and dividing by 1000. -->
<!-- New in Revision 1 -->
## Modularity estimates

@@

<!-- New in Revision 1 -->
**Additional File 9. Bonferroni-corrected Wilcoxon p-values comparing the internal and external degree distributions of phenotype-specific networks having high (>0.85) and low (<0.58) normalized coherence.** Each sheet corresponds to the type of degree distribution (internal and external), and the type of PPI database (STRING and STRINGfilt).
<!-- End New in Revision 1 -->
<!--
# Supplementary material

@@

<!-- New in Revision 1 -->
**Figure 3. Degree distributions differ for high/low-coherence networks.** Internal and external degree distributions for high- and low-coherence networks and size-matched random networks. STRING and STRING Filtered indicate the corresponding PPI database. Bonferroni-corrected Wilcoxon p-values comparing the distributions are in Additional File 9.
<!-- End New in Revision 1 -->

**Figure 4. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).

**Figure 5. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).
